Using reference models in variable selection \vspace{.1in}

### 2020-04-27 paper/ref_approach.tex
minor cleaning

\begin{document}
\maketitle
\begin{abstract}
  Variable selection, or more generally, model reduction is an important aspect of the statistical workflow aiming to provide insights from data. In this paper we discuss and demonstrate the benefits of using a reference model in variable selection. A reference model acts as a noise-filter on the target variable by modeling its data generating mechanism. As a result, using the reference model predictions in the model selection procedure reduces the variability and improves stability leading to improved model selection performance. Assuming that a Bayesian reference model describes the true distribution of future data well, the theoretically preferred usage of the reference model is to project its predictive distribution to a reduced model leading to projection predictive variable selection approach. Alternatively, reference models may also be used in an ad-hoc manner in combination with common variable selection methods.  In several numerical experiments, we investigate the performance of the projective prediction approach as well as alternative variable selection methods with and without reference models. Our results indicate that the use of reference models generally translates into better and more stable variable selection. Additionally, we demonstrate that the projection predictive approach shows superior performance as compared to alternative variable selection methods independently of whether or not they use reference models.
% These benefits are also illustrated on a real-world example of predicting the amount of body fat for which we compare a reference model based selection method to a purely data driven approach.
\end{abstract}

@@

In data based approaches, such as Lasso selection
\citep{tibshirani1996regression} or stepwise backward/forward
regression \citep{venables2013modern,harrell2015regression}, the
observed empirical data distribution is utilised as a proxy of future
data usually in combination with cross-validation or information
criteria to provide estimates for out-of-sample predictive
performance.  In contrast, reference model based methods approximate
the future data generation mechanism using the predictive distribution
of a reference model, which can be, for example, a full-encompassing
model including all variables.  The main assumption under any
reference model approach is that we operate in an
$\mathcal{M}$-\textit{complete} framework
\citep{book:bernardo_smith,vehtari2012survey}, that is, we have
constructed a model which reflects our beliefs about
the future data in the best possible way and which has passed model checking and criticism
\citep[see, e.g.][]{gelman2013bayesian,gabry2019visualization}.  The reference model approach
has been used in Bayesian statistics at least since the seminal
work of \citet{paper:reference_lindley}. For more historical
references, see \citet{vehtari2012survey} and
\citet{paper:model_selection}, and for most recent methodological
developments see \citet{paper:projpred}.

@@

Here we repeat the selection of Section \ref{bodyfat-1} via
stepwise backward regression. In this case, the overall number of
variables (original plus noisy) is 100, as it was in the last part of
Section \ref{bodyfat-1}. We compare results with and without using
a simple reference model approach outlined in \eqref{eq:ref_mod} with
steplm. Figure \ref{fig:bodyfat_step_refvsdata} shows the number of
irrelevant variables included in the final model and the out-of-sample
root mean square error (RMSE). Results are based on 100 bootstrap
samples on the whole dataset, and the predictive performance is tested
on the observations excluded at each bootstrap sample. We observe that
the reference model reduces the number of irrelevant variables included
in the final model. This leads to less overfitting and thus to
improved out-of-sample predictive performance in terms of RMSE. The
reference model approach applied to the stepwise backward regression
achieves outstanding improvements considering its simplicity, yet it
does not reach the goodness of the much more sophisticated projective
prediction approach (see results of Section \ref{bodyfat-1}).
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_step_refvsdata.pdf}
  \vspace{-0.7\baselineskip}
  \caption{Body fat example: Stepwise backward selection with and without using a reference model. The x-axis denotes the number of selected irrelevant variables on the left and the out-of-sample RMSE on the right-hand side based on 100 bootstrap samples. The reference approach reduces the number of noisy variables selected and the out-of-sample RMSE.}
  \label{fig:bodyfat_step_refvsdata}
\end{figure}
\begin{table}[tp]
\footnotesize
\centering
\begin{tabular}{l|rr}
%  \hline
Method & RMSE & Selected noisy \\ 
  \hline
Stepwise selection & 6.9 (0.7) & 48 (7)    \\
Reference model + Stepwise selection & 4.7 (0.5) & 25 (7)   \\
\end{tabular}
\caption{Body fat example: Mean and standard deviation (between brackets) values of Figure \ref{fig:bodyfat_step_refvsdata}.}
\label{tab:step_refvsdata}
\end{table}

>>>>

\begin{document}
\maketitle
\begin{abstract}
  Variable selection, or more generally, model reduction is an important aspect of the statistical workflow aiming to provide insights from data. In this paper, we discuss and demonstrate the benefits of using a reference model in variable selection. A reference model acts as a noise-filter on the target variable by modeling its data generating mechanism. As a result, using the reference model predictions in the model selection procedure reduces the variability and improves stability leading to improved model selection performance. Assuming that a Bayesian reference model describes the true distribution of future data well, the theoretically preferred usage of the reference model is to project its predictive distribution to a reduced model leading to projection predictive variable selection approach. Alternatively, reference models may also be used in an ad-hoc manner in combination with common variable selection methods.  In several numerical experiments, we investigate the performance of the projective prediction approach as well as alternative variable selection methods with and without reference models. Our results indicate that the use of reference models generally translates into better and more stable variable selection. Additionally, we demonstrate that the projection predictive approach shows superior performance as compared to alternative variable selection methods independently of whether or not they use reference models.
% These benefits are also illustrated on a real-world example of predicting the amount of body fat for which we compare a reference model based selection method to a purely data driven approach.
\end{abstract}

@@

In data-based approaches, such as Lasso selection
\citep{tibshirani1996regression} or stepwise backward/forward
regression \citep{venables2013modern,harrell2015regression}, the
observed empirical data distribution is utilised as a proxy of future
data usually in combination with cross-validation or information
criteria to provide estimates for out-of-sample predictive
performance.  In contrast, reference model based methods approximate
the future data generation mechanism using the predictive distribution
of a reference model, which can be, for example, a full-encompassing
model including all variables.  The main assumption under any
reference model approach is that we operate in an
$\mathcal{M}$-\textit{complete} framework
\citep{book:bernardo_smith,vehtari2012survey}, that is, we have
constructed a model which reflects our beliefs about
the future data in the best possible way and which has passed model checking and criticism
\citep[see, e.g.][]{gelman2013bayesian,gabry2019visualization}.  The reference model approach
has been used in Bayesian statistics at least since the seminal
work of \citet{paper:reference_lindley}. For more historical
references, see \citet{vehtari2012survey} and
\citet{paper:model_selection}, and for most recent methodological
developments see \citet{paper:projpred}.

@@

Here we repeat the selection of Section \ref{bodyfat-1} via
stepwise backward regression. In this case, the overall number of
variables (original plus noisy) is 100, as it was in the last part of
Section \ref{bodyfat-1}. We compare results with and without using
a simple reference model approach outlined in \eqref{eq:ref_mod} with
steplm. Figure \ref{fig:bodyfat_step_refvsdata} shows the number of
irrelevant variables included in the final model and the out-of-sample
root mean square error (RMSE). Results are based on 100 bootstrap
samples on the whole dataset, and the predictive performance is tested
on the observations excluded at each bootstrap sample. We observe that
the reference model reduces the number of irrelevant variables included
in the final model. This leads to less overfitting and thus to
improved out-of-sample predictive performance in terms of RMSE. The
reference model approach applied to the stepwise backward regression
achieves outstanding improvements considering its simplicity, yet it
does not reach the goodness of the much more sophisticated projective
prediction approach (see results of Section \ref{bodyfat-1}).
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_step_refvsdata.pdf}
  \vspace{-0.7\baselineskip}
  \caption{Body fat example: Stepwise backward selection with and without using a reference model. The x-axis denotes the number of selected irrelevant variables on the left and the out-of-sample RMSE on the right-hand side based on 100 bootstrap samples. The reference approach reduces the number of noisy variables selected and the out-of-sample RMSE.}
  \label{fig:bodyfat_step_refvsdata}
\end{figure}
\begin{table}[tp]
\footnotesize
\centering
\begin{tabular}{l|rr}
%  \hline
Method & RMSE & Selected noisy \\ 
  \hline
Stepwise selection & 6.9 (0.7) & 48 (7)    \\
Reference model + Stepwise selection & 4.7 (0.5) & 25 (7)   \\
\end{tabular}
\caption{Body fat example: Means and standard deviations (between brackets) of the results shown in Figure \ref{fig:bodyfat_step_refvsdata}.}
\label{tab:step_refvsdata}
\end{table}
